---
title: "Market Efficiency Hypothesis"
author: "Instructor: Menghan Yuan"
date: "Fall 2025"
output: 
  xaringan::moon_reader:
    css: ["theme/xaringan-themer.css", "theme/custom.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: yes
      countIncrementalSlides: no
      ratio: '16:9'
    includes:
      in_header: libs/mathjax.html
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
library(kableExtra)
library(tidyverse)

# Set up knitr
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.retina = 3,
  out.width = "80%"
)

# Custom hook for output control
hook_output_default <- knitr::knit_hooks$get('output')
truncate_to_lines <- function(x, n) {
   if (!is.null(n)) {
      x = unlist(stringr::str_split(x, '
'))
      if (length(x) > n) {
         x = c(head(x, n), '...
')
      }
      x = paste(x, collapse = '
')
   }
   x
}
knitr::knit_hooks$set(output = function(x, options) {
   max.lines <- options$max.lines
   x <- truncate_to_lines(x, max.lines)
   hook_output_default(x, options)
})
```

```{r xaringanExtra-extension, echo=FALSE}
# add search box
xaringanExtra::use_search(show_icon = TRUE, position = "bottom-left")
# enable tile view
xaringanExtra::use_tile_view()
```

class: inverse, center, middle

# Introduction to Market Efficiency

---

## What is Market Efficiency?

.large[
A market in which prices always **"fully reflect"** available information is called .highlight[**efficient**].

Formally, the market is said to be efficient with respect to some .highlight[**information set**].
]

--

### Key Implication:

.large[
Efficiency with respect to an information set implies that it is .highlight[**impossible to make economic profits**] by trading on the basis of that information set.
]

.footnote[
**Economic profits** = "abnormal/superior" returns after adjustment for risk and costs
]

---
.tight-bottom[
## Three Forms of Market Efficiency
]

.pull-left.tight-top.tight-bottom[
### 1. Weak-Form EMH
- Historical market trading data
- E.g., Prices, volumes, past returns


### 2. Semistrong-Form EMH
- Weak-form information **+**  
  &ensp;&ensp; All publicly available information
- E.g., Earnings, dividends, accounting data

### 3. Strong-Form EMH
- Semistrong-form information **+**  
  &nbsp;&nbsp; Insider information
- E.g., Mergers, acquisitions, private info
]

--

.pull-right[
### Direction of Implication:

$$\text{Strong-form EMH} \Longrightarrow$$
$$\text{Semistrong-form EMH} \Longrightarrow$$
$$\text{Weak-form EMH}$$

.center[
**Stronger form ⟹ Weaker form**

**Violation of weaker ⟹ Violation of stronger**
]
]

---

## Violations of Market Efficiency

.center[
```{r EMH-table, echo=FALSE}
tibble(
  `Form of EMH` = c("Weak-Form", "Semistrong-Form", "Strong-Form"),
  `Violation Indicates` = c(
    "Technical analysis or charting can earn superior profits",
    "Fundamental analysis can achieve superior returns", 
    "Corporate insiders can make profitable trades"
  ),
  `Information Set` = c(
    "Historical prices & volumes",
    "All public information",
    "All information (including private)"
  )
) %>%
  kable(format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                font_size = 16,
                full_width = TRUE) %>%
  column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%
  column_spec(2, width = "40%") %>%
  column_spec(3, width = "30%")
```
]

--

### Key Insight

.large[
Perfect efficiency exists only in the **idealized frictionless world**. 

Abnormal returns will exist if there are **costs of gathering and processing information**.

⟹ We measure .highlight[**relative efficiency**]
]


---

class: inverse, center, middle

# Are Market Efficient?

---

##  Overview of EMH Tests

The **Efficient Market Hypothesis (EMH)** posits that asset prices fully reflect available information. 

Tests of market efficiency examine whether investors can systematically earn abnormal returns using available information.

**If prices do not move when the information is revealed**, then the market is efficient with respec to that info.

Test for three forms of EMH:
* **Weak-form efficiency:** Prices reflect all past trading information (historical prices and volumes).
* **Semi-strong form efficiency:** Prices reflect all publicly available information (financial statements, news, etc.).
* **Strong-form efficiency:** Prices reflect all information, including private or insider information.


---

## Weak-Form Tests: Patterns in Stock Returns

**Objective:** Can past price patterns predict future returns?

- **Random Walk Hypothesis:** Stock prices follow a random walk, meaning future price changes are independent of past changes.
  - Cowles and Jones Test for random walk

- **Serial correlation** measures the relationship between current and past returns:
  - Autocorrelation tests (Box-Pierce, Ljung-Box)

- **Indicators of violations of weak-form efficiency:**
    * **Positive serial correlation (momentum):** positive returns tend to follow positive returns.
    * **Negative serial correlation (reversal):** positive returns tend to be followed by negative returns.

---

### Empirical Findings

* Weekly returns on NYSE stocks show **small positive serial correlation** (short-term trends exist but not strong enough for profitable trading).

  * **Interpretation:** weak trends exist, but opportunities for abnormal profits are limited.

* Certain variables predict market returns:

  * Dividend-price ratio (Dividend yield)
  * Earnings-price ratio (Earnings yield)
  * Bond yield spreads (high vs low grade)

  **Interpretation:**

  * Predictability may reflect **time-varying market risk premiums**, not mispricing.
  * Using these variables does **not necessarily violate EMH**.


---
.tight-bottom[
## Semi-Strong Form Tests: Market Anomalies
]
- **Objective:** Can publicly available information beyond past prices generate abnormal returns?
- **Test:** Whether financial ratios or firm characteristics predict excess returns.
- Examples of anomalies:
  1. **P/E ratio effect:** low P/E stocks historically earn higher returns than high P/E stocks.
     
     * Interpretation: may reflect **risk not captured by CAPM**, not true mispricing.
  2. **Size effect:** small-cap stocks earn higher returns than large-cap stocks, even after risk adjustment.
     
     * Possible explanation: **higher risk**, less coverage, or **neglected-firm effect**.
  3. **Book-to-market effect:** high book-to-market stocks (value stocks) earn higher average returns than low book-to-market (growth) stocks.

- **Risk adjustment is critical:** using CAPM or other risk models (e.g., FF 3-factor) may change whether an apparent anomaly violates EMH.

---

### Example: Small-Firm Effect

```{r small-firm-effect, echo=FALSE, fig.height=3, fig.asp=0.5}
AR <- tibble(
    size = seq(1, 10),
    annual_return = c(18.37, 16.55, 15.90, 15.54, 14.73, 14.50, 14.12, 13.52, 12.63, 11.05)
)
ggplot(AR, aes(x = size, y = annual_return)) +
    geom_bar(stat = "identity", fill = "#3498DB", width = 0.6) +
    geom_text(aes(label = sprintf("%.2f", annual_return)),
        vjust = -0.5, size = 4
    ) + # adjust vjust for position
    labs(
        title = "Average annual return for 10 size-based portfolios, 1926–2018",
        x = "Size Decile (1=Smallest; 10=Largest)",
        y = "Average Annual Return (%)"
    ) +
    scale_y_continuous(limits = c(0, 20), breaks = seq(0, 20, by = 2)) +
    theme_minimal() +
    theme(text = element_text(size = 12))
```


---

.tight-bottom[
### The Small-Firm Premium
]
- **7.32%** annual return difference: Small firms vs. Large firms
- Premium persists even after CAPM risk adjustment
- Simple rule: "invest in small-cap stocks" → excess returns

--

.tight-bottom[
### The Puzzle
]
- Any investor can measure firm size at **minimal cost**
- Why should such simple strategy yield **large rewards**?

--

.tight-bottom[
### Main Takeaway:
]
- **Alternative explanations**:
  - **CAPM limitations**: May not capture all risk factors for small firms

     Higher true risk (not captured by mkt beta) → indicate necessity of FF size factor
  - Less analyst coverage  
  - Neglected-firm effect
- **Not necessarily** market inefficiency

---

## Testing Market Efficiency: The Challenge

### **The Joint Test Problem**

Any test of efficiency must assume an equilibrium model that defines normal security returns.  
$\Rightarrow$ Our null hypothesis is a .highlight[**joint test**] of:

- Market efficiency **AND** 
- Correct equilibrium model

When we reject the null, this could be because:
- The market is truly inefficient, **OR**
- An incorrect equilibrium model has been assumed

---

### Joint Test Problem: Consequences

- The result of this is that when we reject the null, this could be because the market is truly inefficient or because an incorrect equilibrium model has been assumed.

- With the wrong asset pricing model, we can wind up rejecting efficiency. It would be easy to find abnormal returns to be forecastable if we had the wrong equilibrium.

- Asset pricing models give the equilibrium return; abnormal returns are deviations of real prices from equilibriums.

- **Key takeaway:** Semi-strong tests examine **public information** but interpretation depends heavily on **risk adjustment methodology**.

---

## Strong-Form Tests

* **Objective:** Do prices reflect all information, including insider knowledge?
* **Tests:** Examine returns of insiders or corporate insiders who have non-public information.
  * If markets are strong-form efficient, insiders should not earn abnormal returns.
* **Findings:**

  * Insiders often earn abnormal profits, indicating **markets are not perfectly strong-form efficient**. 
    
     → Strong form is generally **rejected in practice**.
  * Can outsiders exploit this by mimicking insider trades?
    
     **No, after costs.** Insider trades must be reported to the SEC and become public information. Abnormal returns are small and disappear once transaction costs are considered.

---
### Implications for EMH

<div style="margin-top: 5rem;">

- **Markets are not strong-form efficient** (insiders have private information advantage).

- But markets are **close to semi-strong efficient**: once information is public, it is quickly incorporated, leaving no free profit opportunities for ordinary investors.


---

class: inverse, center, middle

# Random Walk as a Test of EMH

---

## Random Walk and Efficient Markets

- EMH says: stock prices already reflect all available information.
  - Weak-form: past prices
  - Semi-strong: all public info
  - Strong-form: all info (private and public)

- If prices are predictable → some information is *not* yet reflected → EMH violated.

- Therefore:

    - **Evidence of predictability** = evidence of inefficiency.

    - **Evidence of randomness** = consistent with market efficiency.

- How Random Walk Tests Relate to Market Efficiency
  - Cannot reject Random Walk ⟺ Weak-form EMH holds
  - Reject Random Walk ⟺ Weak-form EMH fails

---
## Random Walk Process

.large[
A process $\{p_t\}$ is called a .highlight[**Random Walk**] (RW) if:

$$p_t = \mu + p_{t-1} + e_t$$

where $E(e_t)=0$ and $\text{Var}(e_t)=\sigma^2$.
]


### Components:
- $\red{p_t}$: natural logarithm of asset price at time $t$, $p_t = \log P_t.$
  
  Note that $r_t = p_t - p_{t-1}$ is the **log return**, and can be expressed as:
  $$r_t = \mu + e_t .$$
- $\red{\mu}$: Expected price change or **drift**
- $\red{e_t}$: Innovations, disturbances, or **increments**

---
.tight-bottom[
## Three Forms of Random Walk
]

.pull-left.tight-top.tight-bottom[
.small-text.red[
### RW1: IID Increments
]

$$e_t\sim \text{IID } (0, \sigma^2)$$

- Independently and Identically Distributed
- **Strongest** form

.small-text[
### RW2: INID Increments
]

$$e_t\sim \text{INID } (0, \sigma^2)$$

- Independent but Not necessarily Identical
- Allows for **heteroskedasticity**
- Weaker than RW1
]

.pull-right[

.small-text[
### RW3: Uncorrelated
]
$$\text{Cov}[e_t, e_{t-k}]=0  \text{ for all } k \neq 0$$
- **Weakest** form
- Allows for dependence in higher moments
]

---
## Cowles and Jones Test (1937)

.large[
Tests RW1 using the frequency of .highlight[**sequences**] and .highlight[**reversals**] in historical stock returns.
]

--
.tight-top.tight-bottom[
### Definitions:
]

- **Sequences**: Consecutive returns with **same sign**
  
  $$(+,+) \text{ or } (-,-)$$

- **Reversals**: Consecutive returns with **opposite signs**
  
  $$(+,-) \text{ or } (-,+)$$

- Indicator Variable:

$$I_t = \begin{cases}
1 & \text{if } r_t>0 \\
0 & \text{if } r_t\leq0
\end{cases}$$


---

## Sequences and Reversals: Formal Definition

### Number of Sequences and Reversals:

.pull-left[
**Number of Sequences:**
$$N_s \equiv \sum_{t=1}^n Y_t$$

**Sequence indicator:**
$$\red{Y_t} \equiv I_tI_{t-1} + (1-I_t)(1-I_{t-1})$$

**Number of Reversals:**
$$N_r \equiv n-N_s$$

Note that $N_s + N_r = n,$ total number of observations.
]

.pull-right[
**Truth Table for $Y_t$:**

| $I_{t-1}$ | $I_t$ | $Y_t$ |
|:---------:|:-----:|:-----:|
| 0 | 0 | 1 |
| 0 | 1 | 0 |
| 1 | 0 | 0 |
| 1 | 1 | 1 |

That is,
$$Y_t = \begin{cases}
1 & \text{if sequence} \\
0 & \text{if reversal}
\end{cases}$$
]

---
.tight-bottom[
## Random Walk without Drift
]
We first consider a driftless RW, where $\mu=0$: $r_t=p_t-p_{t-1}=e_t$

--
.tight-top.tight-bottom.small-text[
### Under Driftless RW:
]
- $P(r_t>0)=\frac{1}{2},$ that is, it is equally likely to have a positive or negative return on any given day.
- For any pair of consecutive returns, a **sequence** and a **reversal** are .highlight[**equally probable**].  
  ⟹ We expect $N_s \approx N_r$


--
.tight-top.tight-bottom.small-text[
### Cowles-Jones Test Statistic:
]
$$\widehat{\text{CJ}} \equiv \frac{N_s}{N_r} = \frac{N_s/n}{N_r/n} = \frac{\hat{\pi}_s}{1-\hat{\pi}_s}$$
where $\red{\pi_s}$ is the **probability of a sequence**; and $1-\pi_s$ is the probability of a reversal.

--
.tight-top.tight-bottom.small-text[
### Under RW Null:
]
$$\widehat{\text{CJ}} \xrightarrow{p} \frac{\pi_s}{1-\pi_s} = \frac{1/2}{1-1/2} = 1$$

---
.tight-bottom.tight-top[
## Random Walk with Drift
]

Suppose log prices follow a RW with drift:
$$p_t=\mu+p_{t-1}+e_t, \quad e_t\sim N(0, \sigma^2)$$
$r_t=\mu+e_t$ where $p_t=\mu+p_{t-1}+e_t$ and $e_t\sim N(0, \sigma^2)$

Such a drift makes **sequences more likely than reversals**.
The indicator variables $I_t$ is like a biased coin-toss:

$$I_t = \begin{cases}
1 & \text{with prob. } p= \color{#b94047}{\pi} \\
0 & \text{with prob. } 1-p= 1-\pi
\end{cases}$$

where $\pi$ is the probability of a positive return.
$$\pi \equiv P(r_t>0) = \Phi \left(\frac{\mu}{\sigma}\right)$$


--
.small-text.tight-top.tight-bottom[
### Drift Effects:
]
- If $\mu=0$: $\pi = \frac{1}{2}$ (fair coin)
- If $\mu>0$: $\pi > \frac{1}{2}$ (upward bias)  
- If $\mu<0$: $\pi < \frac{1}{2}$ (downward bias)

---

## CJ Test with Drift

- Cowles-Jones ratio:
    $$\text{CJ} = \frac{\pi_s}{1-\pi_s} = \frac{\pi^2+(1-\pi)^2}{2\pi(1-\pi)}$$

    It is normally distributed:

    $$\widehat{\text{CJ}} \overset{a}{\sim} N \left(\frac{\pi_s}{1-\pi_s}, \frac{\pi_s(1-\pi_s) + 2\big[\pi^3+(1-\pi)^3-\pi_s^3\big]}{n(1-\pi_s)^4} \right)$$

--
- Test Hypothesis:
  - $H_0$: $\widehat{\text{CJ}}=1$ (Random Walk)
  - $H_1$: $\widehat{\text{CJ}} \neq 1$ (Not Random Walk)

--
- Rejection rule
  - Reject $H_0$ if $|\frac{\widehat{\text{CJ}}-1}{\sqrt{\text{Var}(\widehat{\text{CJ}})}}| > z_{\alpha/2}$

---
.tight-bottom[
## CJ Test: Example
]
```{r out.width="70%"}
f_name <- "images/RW.png"
include_graphics(f_name)
```


---

## CJ Test: Exercise


**Background Story:**

Elena Martinez, a quantitative analyst at Nordic Capital, has been tasked with evaluating the efficiency of TechStart Index, a new technology stock index launched 100 trading days ago. Her supervisor believes that this young market segment might not follow a random walk due to:

- High volatility from speculative trading
- Limited analyst coverage of small tech companies
- Momentum trading by retail investors

Elena decides to use the classic Cowles-Jones test to determine if the daily returns show any predictable patterns that could be exploited for trading profits.

---

**The Data:**

Elena collected 100 daily log returns from the TechStart Index. After analyzing the sign patterns of consecutive returns, she found:

- **48 sequences** (consecutive days with returns of the same sign: $(+,+)$ or $(-,-)$)
- **51 reversals** (consecutive days with opposite signs: $(+,-)$ or $(-,+)$)
- **56 positive return days** out of 100 total days


**Your Task:**

As Elena's junior analyst, help her complete the Cowles-Jones test analysis:

### **Questions:**

**Q1.** What is the Cowles-Jones test statistic $\widehat{CJ}$ for this data?

**Q2.** Under the random walk hypothesis (no drift), what should the expected value of $\widehat{CJ}$ be? Explain why.

**Q3.** Calculate the proportion of positive returns ($\pi$) and the proportion of sequences ($\pi_s$). What do these values suggest about the market?

---

**Q4.** Calculate the standard deviation $\sigma$ for the Cowles-Jones test statistic using the formula:
$$\sigma^2 = \frac{\pi_s (1-\pi_s) + 2(\pi^3 + (1-\pi)^3 - \pi_s^2)}{n (1-\pi_s)^4}$$
Show all your work step by step.

**Q5.** Using your calculated standard deviation from Q4, compute the t-statistic for testing $H_0: CJ = 1$.

**Q6.** With the t-statistic obtained in Q5, what is the two-sided p-value? (Use standard normal table)

**Q7.** At the 5% significance level, what should Elena conclude about market efficiency in the TechStart Index? Should Nordic Capital develop a trading strategy based on this pattern?

**Q8.** **Business Decision:** Elena's supervisor asks: "Should we allocate $10 million to a momentum-based trading strategy for TechStart stocks?" Provide a recommendation based on your statistical analysis.


---

## Serial Correlation and Random Walk

### Under Random Walk Hypothesis:

Increments $e_t$ are .highlight[**uncorrelated**]:

$$\gamma(k) \equiv \text{Cov}(e_t, e_{t-k}) = 0 \quad \text{for all } k\neq 0$$

Or equivalently:

$$\rho(k) \equiv \frac{\text{Cov}(e_t, e_{t-k})}{\sqrt{\text{Var}(e_t)\text{Var}(e_{t-k})}} = 0 \quad \text{for all } k\neq 0$$


--

### Key Insight:
.large[
$$\text{Cov}(e_t, e_{t-k}) = 0 \quad \Leftrightarrow \quad \text{Cov}(r_t, r_{t-k}) = 0$$

The covariance of **first-differences** equals the covariance of **increments**.
]

---

## Sample Autocorrelations

### Sample Estimates:
For a given sample $\{r_t\}_{t=1}^T$:

$$\begin{aligned}
\hat{\gamma}(k) & =  \frac{1}{T} \sum_{t=1}^{T-k} (r_t-\bar{r}_T) (r_{t+k}-\bar{r}_T), \quad 0\leq k <T \\
\hat{\rho}(k) &= \frac{\hat{\gamma}(k)}{\hat{\gamma}(0)}
\end{aligned}$$

where $\bar{r}_T \equiv \frac{1}{T} \sum_{t=1}^{T} r_t$

--

Sample autocorrelation coefficients are asymptotically independent and normally distributed:

$$\sqrt{T} \hat{\rho} (k) \overset{a}{\sim} N(0,1)$$

The sampling distribution yields a variety of autocorrelation-based tests.

---

## Box-Pierce Q-Statistic

.tight-top.tight-bottom[
### Joint Hypothesis Test:
]

Tests whether the first $m$ autocorrelations are **jointly zero**:

$$H_0: \rho_1 = \rho_2 = \cdots = \rho_m = 0.$$

--

.tight-bottom[
### Box-Pierce Q-Statistic (1970):
]

$$Q_m \equiv T\sum_{k=1}^m \hat{\rho}^2(k) \sim \chi_m^2$$

.tight-top.tight-bottom[
### Ljung-Box Correction (1979):
]

$$Q_m' \equiv T(T+2)\sum_{k=1}^m \frac{\hat{\rho}^2(k)}{T-k} \sim \chi_m^2$$

.footnote[
The Ljung-Box version provides better fit to $\chi_m^2$ distribution for small samples
]

---
.tight-bottom[
## Box-Pierce Test: Application
]
```{r}
set.seed(123)
t <- seq(1, 100, by=1)
et <- rnorm(length(t), mean=0, sd=1) # IID N(0,1) increments
pt <- c(0, cumsum(et))  # driftless random walk
```

```{r box-pierce-test, echo=TRUE}
returns <- diff(pt)  # First differences = returns
# Box-Pierce test  
Box.test(returns, lag=1, type="Box-Pierce")
```

```{r ljung-box-test, echo=TRUE}
# Ljung-Box test
Box.test(returns, lag=1, type="Ljung-Box")
```

**Interpretation**: High p-values suggest we cannot reject the null hypothesis of zero autocorrelation, consistent with random walk.

---
## Variance Ratio Test

- The VR(q) Statistic:
    $$\text{VR(q)} \equiv \frac{\text{Var}[r_t(q)]}{q\cdot \text{Var}[r_t]}$$
    where $r_t(q)\equiv \sum_{k=0}^{q-1} r_{t-k}$ is the **q-period return**
--

- Under Zero Autocorrelation:
    $$\text{Var}[r_t(q)] = q\cdot \text{Var}[r_t]$$
    Therefore: $$\text{VR(q)}=1$$

--

- When Serial Correlation Exists:
    $\text{VR(q)}$ moves away from $1$:

  - **Positive serial correlation** ⟹ $\text{VR(q)} > 1$
  - **Negative serial correlation** ⟹ $\text{VR(q)} < 1$


---

## Summary: Testing Market Efficiency

.tight-top.tight-bottom.small-text[
### Three Main Approaches:
]

.pull-left.tight-top.tight-bottom.small-text[
#### 1. **Cowles-Jones Test**
.tight-top.tight-bottom[
- Tests using sequences/reversals
- Simple intuitive approach
- Works for basic RW testing
]


#### 2. **Autocorrelation Tests**
.tight-top.tight-bottom[
- Direct test of serial correlation
- Box-Pierce/Ljung-Box Q-statistics
- Tests joint hypothesis of zero autocorrelations
]

#### 3. **Variance Ratio Tests**
.tight-top.tight-bottom[
- Compare variance of multi-period vs single-period returns
- Powerful against various forms of dependence
]
]

--

.pull-right.tight-top.tight-bottom[
### .highlight[**Key Insights**]:

- **Joint test problem**: Always testing efficiency AND asset pricing model simultaneously
- **Most tests suggest**: If security returns cannot be forecasted, then market efficiency cannot be rejected
- **Perfect efficiency** exists only in idealized frictionless world
- Focus on **relative efficiency** in practice
]

---

## References

.small[
Box, G. E. P., & Pierce, D. A. (1970). Distribution of residual autocorrelations in autoregressive-integrated moving average time series models. *Journal of the American Statistical Association*, 65(332), 1509-1526.

Campbell, J. Y., Lo, A. W., & MacKinlay, A. C. (1997). *The econometrics of financial markets*. Princeton University Press.

Cowles, A., & Jones, H. E. (1937). Some a posteriori probabilities in stock market action. *Econometrica*, 5(3), 280-294.

Ljung, G. M., & Box, G. E. P. (1978). On a measure of lack of fit in time series models. *Biometrika*, 65(2), 297-303.

Lo, A. W., & MacKinlay, A. C. (1988). Stock market prices do not follow random walks: Evidence from a simple specification test. *The Review of Financial Studies*, 1(1), 41-66.

Lo, A. W. (1989). The size and power of the variance ratio test in finite samples: A Monte Carlo investigation. *Journal of Econometrics*, 40(2), 203-238.
]

